HO unification from object language to meta language

### 2024-02-20 paper/main.tex
tex clean

\newcommand{\U}{\ensuremath{=_o}}
\newcommand{\Ue}{\ensuremath{=_\lambda}}
\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}}}
\newcommand{\Ho}{\ensuremath{\mathcal{H}_o}}

@@

  The object logic we are interested in is Coq's~\cite{Coq-refman}
  Dependent Type Theory (DTT),
  for which we aim to implement a unification procedure \U{} using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi comes equipped with the equational theory \Ue{}, comprising
  $\eta\beta$ equivalence
  and higher order unification restricted to the pattern
  fragment~\cite{miller92jsc}.
  We want \U{} to feature the same equational theory as \Ue{} but on the
  object logic DTT. Elpi also comes with an encoding for DTT that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
  Unfortunately this encoding, which we refer to as \Fo{},
  ``underuses'' \Ue{} by restricting it to first-order unification problems only. 
  To address this issue, we propose a better-behaved encoding, \Ho{},
  demonstrate how to map unification problems in \Fo{}
  to related problems in \Ho{}, and illustrate
  how to map back the unifiers found by \Ue{}, effectively implementing
   \U{} on top of \Ue{} for the encoding \Fo{}.

@@

\noindent
The encoding of DTT provided by Elpi, that we will discuss at length later in
section~\ref{sec:encodings} and ~\ref{} features the following term constructors:

@@

\noindent
This time the root cause is that the unification procedure of \Ue{} of the
meta language is not aware of the equational theory of the object logic \U{},
even if both theories include $\eta\beta$-conversion and admit most general
unifiers for problems in the pattern fragment~\cite{miller92jsc}.

In this paper we discuss alternative encodings of Coq in
Elpi~\ref{sec:encodings}, then we identify a minimal language \Fo{}
in which the problems sketched here can be fully described.
We then detail an encoding \elpiIn{comp} from \Fo{} to \Ho{} (the language of
the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
bla bla..

@@

% Why the fo prefix for fo_tm && Uvar in OL have no scope
Because of our translation of the terms of the object language into the meta
language we have sketched in section one, a object language term has no scope.
An example of this claim can be given by analyzing deeply the subterm \coqIn{P x} from
the instance forall\_dec. \coqIn{P} is an higher order variables
with type \coqIn{A ~$\to$~ Prop} and
\coqIn{x} is a name bound to \coqIn{P}.
However, at the meta level, the translation of \coqIn{P x} become \elpiIn{app[uv
1, x]}, that is, \elpiIn{uv 1} cannot see \elpiIn{x}.

@@

The \eqho predicate tests if two terms are equal in the world of the meta
languages. This equality function is not capable to understand the \etabeta
reductions over terms of the object language. The ML sees a term of the type
\elpiIn{tm} as atom on which only structural equality can be performed. It is
only possible to dereference a variable if assigned.

@@

@@

le due sintassi, XX tradotto, unif passa, bisogna riportare la soluzione
nel mondo fo

this should hold in both encoding
%abella
\begin{elpicode}
forall t1 t2 s, unif t1 t2 nil s -> equal s t1 t2
\end{elpicode}
we have unif only in the HO one, where equal is
weaker (but unif is stronger)

segnatura di decomp e unif, equalfo (fa beta, eta, deref)

@@

equal fo is congruence, deref, beta, eta (for the OL).
equal ho is congruence, deref (for the OL).
unif (fo or ho) should cover pattern fragment, i.e. well behaved unif.

ho, and unif ho, is the LF/ML language/unif. here we describe fo and ho
with a deeply embedded syntax, but the idea is that uv can be mapped to
an actual unif variable.

of course we could write a unif for the ho that had beta and eta,
but if we use the one of the LF it can't have eta beta of the OL baked in.

@@

\begin{elpicode}
prove (app[c "decidable", all x\ P x]) Proof :- ho-link P P', prove (... P' ...).
\end{elpicode}

qui P nel paper diventa uv N [x] per un certo N.

\section{recovering HO}

\begin{elpicode}
  :index (1)
  p (all x\ app[F,x]).
\end{elpicode}

fallisce perche le liste non hanno la stessa lunghezza.

compile

\begin{elpicode}
  p (all x\ F' x) = p (all x\ app[f,x,x]) , link F F'.
  F' = x\ app[f,x,x]
  F = lam a\ app[f,a,a].
\end{elpicode}

\begin{elpicode}
  type comp fo.tm -> ho.tm -> list link -> list link -> ho.subst -> ho.subst -> o.
  comp (fo.c X)   (ho.c X) L L S S.
  comp (fo.app [fo.uv N|Argsss]) TT L L3 S S3 :- %!,
    % TODO: here split-pf to enter dist. names into N
    % split-pf Argsss [] PF NPF,
    split-pf Argsss [] PF [], NPF = [], % TODO: compile to (uv N L) + link-fo-app
    print "In PF" PF NPF, 
    if (NPF = []) (TT = ho.uv M PF1) (TT = ho.app [ho.uv M PF1 | NPF1]),
    % pattern-fragment Args,
    fold4 comp PF PF1 L L1 S S1,
    fold4 comp NPF NPF1 L1 L2 S1 S2,
    ho.new S2 M S3,
    % TODO: maybe len can be given by split-pf
    len PF Len,
    L3 = [link N M Len| L2].
  % TODO: if don't want to modify unif, we compile `fo.app [fo.c f, c0, c0, ho.c a]` into 
  % `ho.app [ho.app[f, c0, c0], ho.c a]
  comp (fo.app A) (ho.app A1) L L1 S S1 :- fold4 comp A A1 L L1 S S1.
  comp (fo.lam F) (ho.lam F1) L L1 S S1 :- 
    (pi x y\ (pi A S\ comp x y A A S S) => comp (F x) (F1 y) L L1 S S1).
  comp (fo.uv N) (ho.uv M []) L [link N M z|L] S S1 :- ho.new S M S1.

  decmop...
\end{elpicode}

>>>>

\newcommand{\U}{\ensuremath{=_o}\xspace}
\newcommand{\Ue}{\ensuremath{=_\lambda}\xspace}
\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}\xspace}}
\newcommand{\Ho}{\ensuremath{\mathcal{H}_o}\xspace}

@@

  The object logic we are interested in is Coq's~\cite{Coq-refman}
  Dependent Type Theory (DTT),
  for which we aim to implement a unification procedure \U using the ML
  Elpi~\cite{dunchev15lpar}, a dialect of $\lambda$Prolog.
  Elpi comes equipped with the equational theory \Ue, comprising
  $\eta\beta$ equivalence
  and higher order unification restricted to the pattern
  fragment~\cite{miller92jsc}.
  We want \U to feature the same equational theory as \Ue but on the
  object logic DTT. Elpi also comes with an encoding for DTT that works well
  for meta-programming~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
  Unfortunately this encoding, which we refer to as \Fo,
  ``underuses'' \Ue by restricting it to first-order unification problems only. 
  To address this issue, we propose a better-behaved encoding, \Ho,
  demonstrate how to map unification problems in \Fo
  to related problems in \Ho, and illustrate
  how to map back the unifiers found by \Ue, effectively implementing
   \U on top of \Ue for the encoding \Fo.

@@

\noindent
The encoding of DTT provided by Elpi, that we will discuss at length later in
\cref{sec:encodings, ...} features the following term constructors:

@@

\noindent
This time the root cause is that the unification procedure of \Ue of the
meta language is not aware of the equational theory of the object logic \U,
even if both theories include $\eta\beta$-conversion and admit most general
unifiers for problems in the pattern fragment~\cite{miller92jsc}.

In this paper we discuss alternative encodings of Coq in
Elpi~\ref{sec:encodings}, then we identify a minimal language \Fo
in which the problems sketched here can be fully described.
We then detail an encoding \elpiIn{comp} from \Fo to \Ho (the language of
the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
bla bla..

@@

% Why the fo prefix for fo_tm && Uvar in OL have no scope
Because of our translation of the terms of the object language into the meta
language we have sketched in section 1, a object language term has no scope.
An example of this claim can be given by analyzing deeply the subterm \coqIn{P x} from
the instance forall\_dec. \coqIn{P} is an higher order variables
with type \coqIn{A ~$\to$~ Prop} and
\coqIn{x} is a name bound to \coqIn{P}.
However, at the meta level, the translation of \coqIn{P x} become \elpiIn{app[uv
1, x]}, that is, \elpiIn{uv 1} cannot see \elpiIn{x}.

@@

The \eqho predicate tests if two terms are equal in the world of the meta
languages. This equality function is not capable to understand the \etabeta
reductions over terms of the object language. The ML sees a term of the type
\elpiIn{tm} as atom on which only structural equality can be performed. It is
only possible to dereference a variable if assigned.
\marginpar{TODO: say HO is same as ML}

@@

- we accept HO unif with PF

@@

@@

@@
